\LARGE{{\bf{Gravitational Wave Signals from Multiple Hidden Sectors} \\ }}

### 2019-10-03 gwnnDraft02.tex
small changes

From the above Lagrangians, the Higgs in sectors with $i \geq 0$ will get a vev given by:
\begin{equation}\label{eqn:vevs}
v^i = \sqrt{-(m_H^2)_i/\lambda_i} \sim \Lambda_H\sqrt{\frac{2i + r}{\lambda N}},
\end{equation}
$\lambda_i$ is the quartic coefficient of the scalar potential and is the same across all sectors, $\lambda_i=\lambda$. This is another way to see how this framework can solve the hierarchy problem: the Higgs vev is parametrically smaller than the cutoff for $N \gg 1$. 
%The parameter $r$ is a way to adjust the tuning of the model: adjusting $r$ away from 1 (which leads to purely uniform spacing) allows us to make a larger gap between our sector and the next one. Within this work, we looked at variants on this sort of tuning including non-uniform distributions and clustered sector effects; these ideas are discussed in Sec. \ref{sec:gw}.   
%
%
%
%
%$N$naturalness is a model framework that looks to solve the hierarchy problem through the introduction of $N$ mutually non-interacting sectors \cite{Arkani-Hamed:2016rle}. The framework itself is very general: the various sectors can possess a wide range of particle content that can be freely selected by the model builder. The one exception to this is that ``our" sector must consist of the Standard Model. For simplicity, the original paper takes all additional new sectors to be copies of the SM (with the same gauge groups and Yukawa structure); however the Higgs mass parameters of each sector are allowed to taken on values distributed between $-\Lambda_H^2$ and $\Lambda^2_H$, with $\Lambda_H$ being the scale that cuts off quadratic divergences.
%
%This construction leads to sectors that are automatically accidentally tuned to the $1/N$ level, leading to a sector with a mass parameter $m_H^2 \sim \Lambda_H^2/N$. The sector with the smallest non-zero vacuum expectation value (vev) as ``our" sector --- the SM sector. It should be noted that the values of $m_H^2$ pass through zero, resulting in effectively 2 types of sectors: ``standard" sectors like our own with a negative Higgs mass parameter squared, possess a vev, and exhibit electroweak symmetry breaking and ``exotic" sectors with a positive Higgs mass parameter squared that thus feature no electroweak symmetry breaking and a zero vev. 
%
%Keeping with the original convention, we write our mass parameter as
%\begin{equation}\label{eqn:massParam}
%\left(m_H^2\right)_i = -\frac{\Lambda_H^2}{N}(2i+r),
%\end{equation}
%where $i$ denotes the sector ($-\frac{N}{2}\leq i \leq \frac{N}{2}$). $i = 0$ is the sector that possess the smallest vev and is thus ``our" SM sector. \textcolor{green}{Write an equation that shows how the vev scales with $i$.} The vev for each standard sector is given explicitly by
%\begin{equation}\label{eqn:vevs}
%v^i = m^i_H/\sqrt{\lambda_i} = \sqrt{\frac{2i + r}{N}}\left(\frac{\Lambda_H}{\lambda}\right),
%\end{equation}
%$\lambda_i$ is the quartic coefficient of the scalar potential and is the same across all sectors. The parameter $r$ is a way to adjust the tuning of the model: adjusting $r$ away from 1 (which leads to purely uniform spacing) allows us to make a larger gap between our sector and the next one. Within this work, we looked at variants on this sort of tuning including non-uniform distributions and clustered sector effects; these ideas are discussed in Sec. \ref{sec:gw}.   
%
%As mentioned above, half of all sectors in this model feature a Higgs mass parameter such that $m_H^2 < 0$. 
The ``standard sectors'' with $i>0$ feature electroweak symmetry breaking just like in the SM, however the vevs scale with the changing mass parameter: $v_i \sim v_{\rm SM}\sqrt{i}$. This means that the masses of the fermions and the $W$ and $Z$ will also increase proportional to $\sqrt{i}$. The consequences of this scaling on the confinement scale of QCD in the $i \geq 1$ sectors is further discussed in Sec. \ref{sec:dQCD}.

The ``exotic sectors'' with $i<0$ provide a radical departure from our own. $m_H^2 > 0$ leads to no vev for the Higgs, and electroweak symmetry is only broken at very low scales due to the phase transition from free quarks to confinement at the QCD scale $\Lambda_{QCD}$~\cite{Susskind:1978ms}, and the masses of the $W$ and $Z$ are comparable to those of QCD resonances. The mass of fundamental fermions are produced via four-fermion interactions generated after integrating out the SU$(2)$ Higgs multiplet. This leads to very light fermions: 
\begin{equation}\label{eqn:exoticFM}
m_f \sim y_f y_t \Lambda_{QCD}^3/(m^2_H)_i \leq 100 \,\ev,
\end{equation}
with $y_f$ representing the Yukawa coupling to fermion $f$. As we will see, the extremely light quarks that appear in these sectors dramatically change the nature of the QCD phase transition --- unlike the SM, the transition is strongly first order. Again, this is further developed in Sec. \ref{sec:dQCD}. Crucially, this results in the production of gravitational waves. This is the physical signature we explore in this paper; the calculation and results are presented in Sec. \ref{sec:gw}.

\section{QCD Phase Transition}
\label{sec:dQCD}
%
%\begin{figure*}[tb]
%\centering
%\begin{minipage}[c]{\textwidth}
%\includegraphics[width=1.0\textwidth]{columbia.png}
%\end{minipage}
%\hfill
%\caption{Two possible scenarios for the Columbia Plot with the up and down quark mass taken to be identical. The ``physical point" is the Standard Model. Plot from \cite{Cuteri:2017zcb}
%}
%\label{fig:columbia}
%\end{figure*}
We now study the nature of the QCD phase transition across the different sectors. Due to the confining nature of QCD, the exact nature of the phase transition is often difficult to ascertain analytically and requires the study of lattice simulations. In the SM, it is known that the phase transition is a crossover and does lead to gravitational wave signals~\cite{Aoki:2006we,Bhattacharya:2014ara}. In the general case with 3 or more colours, the phase transition can be strongly first order in two regimes~\cite{SVETITSKY1982423,Pisarski:1983ms,Panero:2009tv}:
\begin{itemize}
\item Three or more light flavours.
\item No light flavours. 
\end{itemize}
Light means mass small compared to the confinement scale $\Lambda_{QCD}$, but what that means quantitatively is not precisely determined. In the SM, the up and down quarks are light, but the strange is not sufficiently light for an SFOPT.
% The results of these studies on the nature of QCD phase transitions is summarized in the so-called ``Columbia" plot (Fig. \ref{fig:columbia}). As seen in the above figure (and explicitly demonstrated through lattice studies \cite{lattice}), our sector is solidly in the weak-cross over area. In order to observe the strong first order transition required for gravitational wave production, at least some of the additional sectors present must feature quarks in either the heavy quark (pure Yang-Mills) or massless limits.    
%
%In the Yang-Mills limit we see $m_{u,d,s} \rightarrow \infty$. This leads to the presence of a global $Z_3$ centre symmetry that is broken at high temperatures but restored low temperatures. Ultimately, this restoration results in a first order phase transition \cite{SVETITSKY1982423}. On the other end of the spectrum, ``massless" quark theories ($m_q \ll \Lambda_{QCD}$ for all quarks) also contain strong phase transitions due to the breakdown of SU$(N)\,\times$ SU$(N)$ chiral symmetry \cite{Pisarski:1983ms}. Both the Yang-Mills and massless limit arguments can be generalized to SU$(N \geq 3)$: coupling both with the additional restriction that we avoid the conformal window indicates that strong first order phase transitions can occur if we have $n = 0$ (pure Yang-Mills) or $3 \leq n < 4N$ (light quark limit) very light flavours.
For the standard sectors in our setup, the quark masses increase with increasing vev, so for sufficiently large $i$, all the quarks will be heavier than $\Lambda_{QCD}$,\footnote{$\Lambda_{QCD}$ does vary with $i$, but the sensitivity is very weak as we will see below.} and those large $i$ sectors will undergo an SFOPT if they are reheated above the the confinement scale. 
%For $N$naturalness (or other mirror sector models), standard sectors with higher vevs possess more massive quarks; eventually, sectors with large enough vevs push us into the upper-right corner. 
Conversely, exotic sectors with zero vev feature six very light quarks, so \textit{all} the exotic sectors undergo SFOPT at the temperature of QCD confinement.

@@

\subsection{Standard Sectors}
As shown in Eq.~(\ref{eqn:vevs}), for standard sectors with increasing index $i$ the vevs of said sectors increase $v_i\propto \sqrt{i}$. This leads to increasingly heavy particle spectra for higher sectors --- eventually leading to sectors that are essentially pure Yang-Mills that feature strong first order phase transitions. This, of course, begs the question: at what index $i$ do said phase transitions begin?
%The actual boundary of region of interest on the Columbia plot is not precisely known and further lattice simulations are required to determine exactly where in parameter space cross-over transitions end and first order transitions begin. In lieu of such studies, we parameterize our ignorance in the form of a ``slop factor" $\kappa$. Assuming the up and down quarks of each sector to be of roughly the same scale, we determine what sector $i$ features $\kappa m_u^i \approx \Lambda_{QCD}$.
Using the methods outlined in the prior section we determine $\Lambda_{QCD}$ to have a relevant value of 
\begin{equation}\label{eqn:lambda2}
\Lambda^i_{(2)} = (m_s^i m_c^i m_b^i m_t^i)^{2/29}(\Lambda^i_{(6)})^{21/29}
\end{equation} 
at the energy scale we're interested in. $\Lambda^i_{(6)}$ is identical for all sectors and is taken to have a standard model value of $\Lambda^{(6)}_{MS} =(89 \pm 6)\, \mev$~\cite{PhysRevD.98.030001}. Rewriting Eq.~(\ref{eqn:lambda2}) in terms of standard model variables, 
\begin{equation}\label{eqn:lambda2adj}
\Lambda^i_{(2)} = (m_s m_c m_b m_t i^2)^{2/29}(\Lambda_{(6)})^{21/29}.
\end{equation}
where $m_q$ without a superscript is the mass of $q$ in the SM. We take the sector with SFOPT to be the ones when the mass of the up quark, down quark, and QCD phase transition scale are all comparable:
\begin{equation}
m^i_u \sim m_u \sqrt{i} \sim (m_s m_c m_b m_t i^2)^{2/29}(\Lambda_{(6)})^{21/29}.
\end{equation}
This can be solved for $i$:
\begin{equation}\label{eqn:critIndex}
i^c \sim \frac{(m_s m_c m_b m_t)^{4/21}(\Lambda_{(6)})^{2}}{( m_u)^{58/21}} \sim 10^6.
\end{equation}
As we will see in Sec.~\ref{sec:reheat}, in the original $N$naturalness setup~\cite{Arkani-Hamed:2016rle}, the energy dumped into the $i$th sector scales as $i^{-1}$, so there will will not be enough energy in the sectors with $i>i^c$ to see a signature of these phase transitions. 
%This critical index demonstrates that even in the most extreme cases, where first order phase transitions occur with the up quark an order of magnitude less than $\Lambda_{QCD}$, the critical index is still $\sim 1000$. Looking back to the energy density scaling in Eq.~(\ref{eqn:edSS})\textcolor{green}{refer to a particular equation} we see that, within the context of $N$naturalness, the energy density of these sectors is at most $10^{-4}$ \textcolor{green}{why not $10^{-3}$?} that of our sector. As a result, \textcolor{green}{As we will show below?} the gravitational waves generated by the standard large $i$ sectors in $N$naturalness will not generate detectable signatures. 
However, if we move away from the original $N$naturalness reheating mechanism and begin exploring mirror sectors with large vevs and with relative energy densities $\rho_i / \rho_{SM} \sim 10\%$, a possibility allowed by current constraints, we can have sectors with relatively high dark QCD scales that produce detectable gravitational waves. From Eq.~(\ref{eqn:lambda2}) we can determine the confinement scale of an arbitrary mirror sector. If we take Higgs vevs as high as the GUT scale $\sim 10^{16}$ GeV, then we can use Eq.~(\ref{eqn:lambda2adj}) to get confinement scales as high as $\sim 38 \,\gev$.  The signals of this sector and other test cases like it are explored in Sec. \ref{sec:gw}.

>>>>

From the above Lagrangians, the Higgs in sectors with $i \geq 0$ will get a vev given by:
\begin{equation}\label{eqn:vevs}
v^i = \sqrt{-(m_H^2)_i/\lambda_i} \sim \Lambda_H\sqrt{\frac{2i + r}{\lambda N}},
\end{equation}
$\lambda_i$ is the quartic coefficient of the scalar potential and is the same across all sectors, $\lambda_i=\lambda$. This is another way to see how this framework can solve the hierarchy problem: the Higgs vev is parametrically smaller than the cutoff for $N \gg 1$. 
%The parameter $r$ is a way to adjust the tuning of the model: adjusting $r$ away from 1 (which leads to purely uniform spacing) allows us to make a larger gap between our sector and the next one. Within this work, we looked at variants on this sort of tuning including non-uniform distributions and clustered sector effects; these ideas are discussed in Sec. \ref{sec:gw}.   
%
%
%
%
%$N$naturalness is a model framework that looks to solve the hierarchy problem through the introduction of $N$ mutually non-interacting sectors \cite{Arkani-Hamed:2016rle}. The framework itself is very general: the various sectors can possess a wide range of particle content that can be freely selected by the model builder. The one exception to this is that ``our" sector must consist of the Standard Model. For simplicity, the original paper takes all additional new sectors to be copies of the SM (with the same gauge groups and Yukawa structure); however the Higgs mass parameters of each sector are allowed to taken on values distributed between $-\Lambda_H^2$ and $\Lambda^2_H$, with $\Lambda_H$ being the scale that cuts off quadratic divergences.
%
%This construction leads to sectors that are automatically accidentally tuned to the $1/N$ level, leading to a sector with a mass parameter $m_H^2 \sim \Lambda_H^2/N$. The sector with the smallest non-zero vacuum expectation value (vev) as ``our" sector --- the SM sector. It should be noted that the values of $m_H^2$ pass through zero, resulting in effectively 2 types of sectors: ``standard" sectors like our own with a negative Higgs mass parameter squared, possess a vev, and exhibit electroweak symmetry breaking and ``exotic" sectors with a positive Higgs mass parameter squared that thus feature no electroweak symmetry breaking and a zero vev. 
%
%Keeping with the original convention, we write our mass parameter as
%\begin{equation}\label{eqn:massParam}
%\left(m_H^2\right)_i = -\frac{\Lambda_H^2}{N}(2i+r),
%\end{equation}
%where $i$ denotes the sector ($-\frac{N}{2}\leq i \leq \frac{N}{2}$). $i = 0$ is the sector that possess the smallest vev and is thus ``our" SM sector. \textcolor{green}{Write an equation that shows how the vev scales with $i$.} The vev for each standard sector is given explicitly by
%\begin{equation}\label{eqn:vevs}
%v^i = m^i_H/\sqrt{\lambda_i} = \sqrt{\frac{2i + r}{N}}\left(\frac{\Lambda_H}{\lambda}\right),
%\end{equation}
%$\lambda_i$ is the quartic coefficient of the scalar potential and is the same across all sectors. The parameter $r$ is a way to adjust the tuning of the model: adjusting $r$ away from 1 (which leads to purely uniform spacing) allows us to make a larger gap between our sector and the next one. Within this work, we looked at variants on this sort of tuning including non-uniform distributions and clustered sector effects; these ideas are discussed in Sec. \ref{sec:gw}.   
%
%As mentioned above, half of all sectors in this model feature a Higgs mass parameter such that $m_H^2 < 0$. 
The ``standard sectors'' with $i>0$ feature electroweak symmetry breaking just like in the SM, however the vevs scale with the changing mass parameter: $v_i \sim v_{\rm SM}\sqrt{i}$. This means that the masses of the fermions and the $W$ and $Z$ will also increase proportional to $\sqrt{i}$. The consequences of this scaling on the confinement scale of QCD in the $i \geq 1$ sectors is further discussed in Sec.~\ref{sec:dQCD}.

The ``exotic sectors'' with $i<0$ provide a radical departure from our own. $m_H^2 > 0$ leads to no vev for the Higgs, and electroweak symmetry is only broken at very low scales due to the phase transition from free quarks to confinement at the QCD scale $\Lambda_{QCD}$~\cite{Susskind:1978ms}, and the masses of the $W$ and $Z$ are comparable to those of QCD resonances. The mass of fundamental fermions are produced via four-fermion interactions generated after integrating out the SU$(2)$ Higgs multiplet. This leads to very light fermions: 
\begin{equation}\label{eqn:exoticFM}
m_f \sim y_f y_t \Lambda_{QCD}^3/(m^2_H)_i \leq 100 \,\ev,
\end{equation}
with $y_f$ representing the Yukawa coupling to fermion $f$. As we will see, the extremely light quarks that appear in these sectors dramatically change the nature of the QCD phase transition --- unlike the SM, the transition is strongly first order. Again, this is further developed in Sec.~\ref{sec:dQCD}. Crucially, this results in the production of gravitational waves. This is the physical signature we explore in this paper; the calculation and results are presented in Sec.~\ref{sec:gw}.

\section{QCD Phase Transition}
\label{sec:dQCD}
%
%\begin{figure*}[tb]
%\centering
%\begin{minipage}[c]{\textwidth}
%\includegraphics[width=1.0\textwidth]{columbia.png}
%\end{minipage}
%\hfill
%\caption{Two possible scenarios for the Columbia Plot with the up and down quark mass taken to be identical. The ``physical point" is the Standard Model. Plot from \cite{Cuteri:2017zcb}
%}
%\label{fig:columbia}
%\end{figure*}
We now study the nature of the QCD phase transition across the different sectors. Due to the confining nature of QCD, the exact nature of the phase transition is often difficult to ascertain analytically and requires the study of lattice simulations. In the SM, it is known that the phase transition is a crossover and does lead to gravitational wave signals~\cite{Aoki:2006we,Bhattacharya:2014ara}. In the general case with 3 or more colours, the phase transition can be strongly first order in two regimes~\cite{SVETITSKY1982423,Pisarski:1983ms,Panero:2009tv}:
\begin{itemize}
\item Three or more light flavours.
\item No light flavours. 
\end{itemize}

Light indicates a mass small compared to the confinement scale $\Lambda_{QCD}$, but what that means quantitatively is not precisely determined. In the SM, the up and down quarks are light, but the strange is not sufficiently light for an SFOPT.
% The results of these studies on the nature of QCD phase transitions is summarized in the so-called ``Columbia" plot (Fig. \ref{fig:columbia}). As seen in the above figure (and explicitly demonstrated through lattice studies \cite{lattice}), our sector is solidly in the weak-cross over area. In order to observe the strong first order transition required for gravitational wave production, at least some of the additional sectors present must feature quarks in either the heavy quark (pure Yang-Mills) or massless limits.    
%
%In the Yang-Mills limit we see $m_{u,d,s} \rightarrow \infty$. This leads to the presence of a global $Z_3$ centre symmetry that is broken at high temperatures but restored low temperatures. Ultimately, this restoration results in a first order phase transition \cite{SVETITSKY1982423}. On the other end of the spectrum, ``massless" quark theories ($m_q \ll \Lambda_{QCD}$ for all quarks) also contain strong phase transitions due to the breakdown of SU$(N)\,\times$ SU$(N)$ chiral symmetry \cite{Pisarski:1983ms}. Both the Yang-Mills and massless limit arguments can be generalized to SU$(N \geq 3)$: coupling both with the additional restriction that we avoid the conformal window indicates that strong first order phase transitions can occur if we have $n = 0$ (pure Yang-Mills) or $3 \leq n < 4N$ (light quark limit) very light flavours.
For the standard sectors in our setup, the quark masses increase with increasing vev, so for sufficiently large $i$, all the quarks will be heavier than $\Lambda_{QCD}$,\footnote{$\Lambda_{QCD}$ does vary with $i$, but the sensitivity is very weak as we will see below.} and those large $i$ sectors will undergo an SFOPT if they are reheated above the the confinement scale. 
%For $N$naturalness (or other mirror sector models), standard sectors with higher vevs possess more massive quarks; eventually, sectors with large enough vevs push us into the upper-right corner. 
Conversely, exotic sectors with zero vev feature six very light quarks, so \textit{all} the exotic sectors undergo SFOPT at the temperature of QCD confinement.

@@

\subsection{Standard Sectors}
As shown in Eq.~(\ref{eqn:vevs}), for standard sectors with increasing index $i$ the vevs of said sectors increase $v_i\propto \sqrt{i}$. This leads to increasingly heavy particle spectra for higher sectors --- eventually leading to sectors that are essentially pure Yang-Mills that feature strong first order phase transitions. This, of course, prompts the question: at what index $i$ do said phase transitions begin?
%The actual boundary of region of interest on the Columbia plot is not precisely known and further lattice simulations are required to determine exactly where in parameter space cross-over transitions end and first order transitions begin. In lieu of such studies, we parameterize our ignorance in the form of a ``slop factor" $\kappa$. Assuming the up and down quarks of each sector to be of roughly the same scale, we determine what sector $i$ features $\kappa m_u^i \approx \Lambda_{QCD}$.
Using the methods outlined in the prior section we determine $\Lambda_{QCD}$ to have a relevant value of 
\begin{equation}\label{eqn:lambda2}
\Lambda^i_{(2)} = (m_s^i m_c^i m_b^i m_t^i)^{2/29}(\Lambda^i_{(6)})^{21/29}
\end{equation} 
at the energy scale we're interested in. $\Lambda^i_{(6)}$ is identical for all sectors and is taken to have a standard model value of $\Lambda^{(6)}_{MS} =(89 \pm 6)\, \mev$~\cite{PhysRevD.98.030001}. Rewriting Eq.~(\ref{eqn:lambda2}) in terms of standard model variables, 
\begin{equation}\label{eqn:lambda2adj}
\Lambda^i_{(2)} = (m_s m_c m_b m_t i^2)^{2/29}(\Lambda_{(6)})^{21/29}.
\end{equation}
where $m_q$ without a superscript is the mass of $q$ in the SM. We take the sector with SFOPT to be the ones when the mass of the up quark, down quark, and QCD phase transition scale are all comparable:
\begin{equation}
m^i_u \sim m_u \sqrt{i} \sim (m_s m_c m_b m_t i^2)^{2/29}(\Lambda_{(6)})^{21/29}.
\end{equation}
This can be solved for $i$:
\begin{equation}\label{eqn:critIndex}
i^c \sim \frac{(m_s m_c m_b m_t)^{4/21}(\Lambda_{(6)})^{2}}{( m_u)^{58/21}} \sim 10^6.
\end{equation}
As we will see in Sec.~\ref{sec:reheat}, in the original $N$naturalness setup~\cite{Arkani-Hamed:2016rle}, the energy dumped into the $i$th sector scales as $i^{-1}$, so there will will not be enough energy in the sectors with $i>i^c$ to see a signature of these phase transitions. 
%This critical index demonstrates that even in the most extreme cases, where first order phase transitions occur with the up quark an order of magnitude less than $\Lambda_{QCD}$, the critical index is still $\sim 1000$. Looking back to the energy density scaling in Eq.~(\ref{eqn:edSS})\textcolor{green}{refer to a particular equation} we see that, within the context of $N$naturalness, the energy density of these sectors is at most $10^{-4}$ \textcolor{green}{why not $10^{-3}$?} that of our sector. As a result, \textcolor{green}{As we will show below?} the gravitational waves generated by the standard large $i$ sectors in $N$naturalness will not generate detectable signatures. 
However, if we move away from the original $N$naturalness reheating mechanism and begin exploring mirror sectors with large vevs and with relative energy densities $\rho_i / \rho_{SM} \sim 10\%$, a possibility allowed by current constraints, we can have sectors with relatively high dark QCD scales that produce detectable gravitational waves. From Eq.~(\ref{eqn:lambda2}) we can determine the confinement scale of an arbitrary mirror sector. If we take Higgs vevs as high as the GUT scale $\sim 10^{16}$ GeV, then we can use Eq.~(\ref{eqn:lambda2adj}) to get confinement scales as high as $\sim 38 \,\gev$.  The signals of this sector and other test cases like it are explored in Sec.~\ref{sec:gw}.
